---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data

First we download the files if necessary and load the dataset into the data.frame activity

```{r}
if(!file.exists("activity.zip")){
	URL<-"https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
	download.file(URL, destfile = "activity.zip")
	unzip("activity.zip")
}
activity <- read.csv("~/RProgramming/Reproducable Research/RepData_PeerAssessment1/activity.csv")
head(activity)
```

## What is mean total number of steps taken per day?

To get a feeling for the distribution inside the dataset we calculate the sum of steps per day.
```{r}
activity_sum<-aggregate(steps ~ date, activity, FUN = sum)
summary(activity_sum)
```

We can plot this data in a histogram

```{r}
hist(activity_sum$steps, breaks = 20,
     xlab = "Total Number of steps",
     main = "Histogram of total steps per day",
     xaxt = "n")
axis(1, at= c(seq(0, 22000, by = 1000)),
     gap.axis = 2.7)
```

What are the means and medians for the data?
```{r}
activity_mean<-aggregate(steps ~ date, activity, FUN = mean); summary(activity_mean)
activity_median<-aggregate(steps ~ date, activity, FUN = median); summary(activity_median)
```
We can see that the median for each day is 0.
This means more than half of the time intervals of each day have 0 measured steps.

## What is the average daily activity pattern?

Let us look at the average steps per interval in a time plot.
```{r}
activity_interval<-aggregate(steps ~ interval, activity, FUN = mean)
plot(activity_interval$steps ~ activity_interval$interval,
     type = "l",
     xlab = "Interval",
     ylab = "Steps",
     main = "Average steps per interval")
```

So what is the interval with the highest average steps taken?
```{r}
max_interval<- activity_interval[which.max(activity_interval$steps), ]
print(max_interval$interval)
```


## Imputing missing values

First let's see how many missing values the dataset has.
```{r}
sum(is.na(activity$steps))
```

Let us replace each missing value with the mean of the corresponding interval.
```{r}
activity_nona <- activity
naindx <- which(is.na(activity_nona$steps))
for(i in naindx){
	activity_nona$steps[i] <- activity_interval[which(activity_interval$interval == activity_nona$interval[i]), "steps"]
}
```

How does this impact the histogram for the data?

```{r}
activity_nonasum<-aggregate(steps ~ date, activity_nona, FUN = sum)
par(mfrow=c(1,2))
hist(activity_nonasum$steps, breaks = 20,
     xlab = "Total Number of steps",
     main = "Histogram of total steps per day\nimputed NAs",
     xaxt = "n",
     ylim = c(0,20))
axis(1, at= c(seq(0, 22000, by = 1000)))
hist(activity_sum$steps, breaks = 20,
     xlab = "Total Number of steps",
     main = "Histogram of total steps per day",
     xaxt = "n",
     ylim = c(0,20))
axis(1, at= c(seq(0, 22000, by = 1000)))
# reset parameters
par(mfrow=c(1,1))
par(mar = c(4, 4, 4, 4))
```

We can see that we gained a few days of data that were missing before replacing the missing data. All of them fall into the same region on the histogram, that had the highest frequency before our imputing. This might hint towards whole days missing their data as the all get the same number of steps assigned to them by our imputing method.

## Are there differences in activity patterns between weekdays and weekends?

To see if there is any difference between workdays and weekends, we add another factor to the dataset calculated from the date provided.

```{r}
activity$weekday <- ifelse(weekdays(as.POSIXct(activity$date)) == "Saturday" | 
			   weekdays(as.POSIXct(activity$date)) == "Sunday",
			   "weekend", "weekday")
activity$weekday <- as.factor(activity$weekday)

activity_weekday<-aggregate(steps ~ interval + weekday, activity, FUN = mean)
weekday<-activity_weekday[which(activity_weekday$weekday == "weekday"),]
weekend<-activity_weekday[which(activity_weekday$weekday == "weekend"),]
```

We then visualise this data in two time plots
```{r}
par(mfrow=c(2,1))
par(mar = c(0, 4, 4, 4))
plot(weekday$steps ~ weekday$interval, type = "l",
     xaxt = "n", yaxt = "n",
     ylim = c(0,250),
     ylab = "")
title(main = "Average steps taken by intervall \nfor weekdays and weekends")
title(main = "Weekdays", line = -1, adj = 0.1)
axis(4, at = seq(0,250, by=50))
axis(1, labels = FALSE, tck = 0.02)
axis(1, labels = FALSE, tck = -0.02)
par(mar = c(4, 4, 0, 4))
plot(weekend$steps ~ weekend$interval, type = "l",
     ylim = c(0,250),
     xlab = "Interval",
     ylab = "")
title(main = "Weekends", line = -1, adj = 0.1)
# reset parameters
par(mfrow=c(1,1))
par(mar = c(4, 4, 4, 4))
```
